---
title: "Cardiovascular Risks in America: A Data- Driven Analysis of Risk Clusters of the Disease Prevalence."
author: "Juan Bol"
format: html
editor: visual
---

<https://github.com/boljuan/FinalAssignment_DataExploration>

### **Summary**

This preliminary analysis explores the prevalence of three major cardiovascular risk factors; high blood pressure, diagnosed diabetes and coronary heart disease using the **2024 CDC PLACES dataset** and the **U.S Department of Agriculture’s Economic Research Service Median Household Income dataset** to measure household income correlation. Cardiovascular disease remains the leading cause of death in the United States, and these three conditions contribute significantly to long-term morbidity and mortality. Understanding their geographic distribution can help the US National Cardiovascular Society identify priority areas for prevention and intervention. Analyzing available data, we have identified national hotspots and conducted a cross-statistical analysis with States where the burden is highest and lowest. The key findings indicate a strong overlap of risk in the rural areas of the Southeast States counties with low household income that suggest inacessibility to preventative care. Our finding lays the foundation for deeper state to even county specific investigation and confident that this analysis supports the Society’s ongoing effort to help address cardiovascular disease and health disparities.

### **Background and Policy Relevance**

According to the Center for Disease Control (CDC) cardiovascular disease remains the leading cause of death in the United States with one in every five deaths imposing an annual economic burden of more than \$400 billion (Centers for Disease Control and Prevention). Despite decades of clinical progress, high blood pressure and diabetes are underdiagnosed and poorly controlled in many regions, particularly in under-served, rural, and low-income counties and racial and ethnic minorities. This analysis aligns with several national objectives of the Center for Disease Control and Prevention (CDC) including [Healthy People 2030](https://odphp.health.gov/healthypeople), [CDC’s Million Hearts Initiative](https://millionhearts.hhs.gov/about-million-hearts/index.html)and the[Affordable Care Act](https://www.hhs.gov/healthcare/about-the-aca/index.html) which is designed to reduce high-cost, high-need cases.

### **Data Source**

The data set used for this analysis came from the Center for Disease Control and Prevention 2024 CDC [PLACES dataset](https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-County-Data-20/swc5-untb/about_data) and the [U.S Department of Agriculture’s Economic Research Service Median Household Income dataset.](https://www.ers.usda.gov/data-products/county-level-data-sets/county-level-data-sets-download-data) This publicly available dataset provides a wide range of model-based estimates.  This analysis focuses on three key selected variables from the dataset that includes the prevalence of high blood pressure, diagnosed diabetics and coronary heart disease among adults at the county level.

These selected variables can be identified within the dataset which has Health Category and Measures variable at county level across 50 states and the District of Columbia. The U.S Department of Agriculture’s Economic Research Service Median Household Income dataset provides median household income for which the year 2022 was used in this analysis.

```{r}
library(dplyr)
library(readr)

#Dataset
file_path <- "PLACES__Local_Data_for_Better_Health__County_Data_2024_release_20250430.csv"
places <- read_csv(file_path)


#data frame of column names and data types
column_info <- data.frame(
  Column_Name = names(places),
  Data_Type = sapply(places, class),
  stringsAsFactors = FALSE
)

print(column_info)

# Save the column information to a CSV file
write_csv(column_info, "places_column_datatype_info.csv")
```

```{r}
library(dplyr)
library(readr)
library(tidyr)

#Dataset
file_path <- "PLACES__Local_Data_for_Better_Health__County_Data_2024_release_20250430.csv"
places <- read_csv(file_path)

#df for Category and Measure pairs
unique_vars <- places %>%
  select(Category, Measure) %>%
  distinct() %>%
  arrange(Category, Measure)

print(unique_vars)
write_csv(unique_vars, "category_measure_df.csv")
```

### **Methods**

For this report, a comprehensive analytical approach was used to uncover spatial and statistical patterns in cardiovascular health outcomes across U.S. counties. Data cleaning in R and ensuring data quality served as the foundational step, ensuring the dataset was standardized and structured for accurate analysis. This was followed by prevalence mapping and clustering, which visualized geographic hotspots and grouped counties based on similar disease burden profiles. Finally, a correlation matrix and co-prevalence detection were conducted to identify statistically significant associations among chronic conditions in a State with the highest prevalence to the lowest. Highlighting regions with overlapping health vulnerabilities was key to share public health approaches recommendations.

### **Findings**

#### **1.1 Cardiovascular Disease Distribution**

Figure 1 illustrates a substantial interstate disparity in high blood pressure, diabetes, and coronary heart disease across the U.S. States. The Southeast states exhibit consistently higher mean prevalence rates for all three conditions, suggesting regional clustering that can be influenced by underlying socio-demographic and environmental factors. This offers critical insight into the geographic distribution of cardiovascular diseases which often share common risk outliners that including poor diet, sedentary lifestyle, limited access to primary care and even environmental conditions.

```{r}
library(dplyr)
library(stringr)
library(ggplot2)
library(sf)
library(tigris)
library(tidyr)
library(scales)
library(readr)

options(tigris_use_cache = TRUE)

# Load datasets
places <- read_csv("PLACES__Local_Data_for_Better_Health__County_Data_2024_release_20250430.csv")
fips_master <- read_csv("county_fips_master.csv")

# Clean and filter
places_clean <- places %>%
  mutate(across(where(is.character), str_squish)) %>%
  filter(Measure %in% c(
    "High blood pressure among adults",
    "Diagnosed diabetes among adults",
    "Coronary heart disease among adults"
  )) %>%
  distinct() %>%
  drop_na(Measure, Data_Value, StateDesc, LocationName) %>%
  mutate(
    StateDesc = str_to_lower(StateDesc),
    LocationName = str_to_lower(LocationName)
  )

fips_master_clean <- fips_master %>%
  mutate(
    state_name = str_to_lower(str_squish(state_name)),
    county_name = str_to_lower(str_squish(county_name))
  )

# Merge datasets
places_with_fips <- places_clean %>%
  left_join(fips_master_clean, by = c("StateDesc" = "state_name", "LocationName" = "county_name"))

# Summarize by state
state_summary <- places_with_fips %>%
  group_by(StateDesc, Measure) %>%
  summarize(mean_value = mean(Data_Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = Measure,
    values_from = mean_value
  ) %>%
  rename(
    high_bp = `High blood pressure among adults`,
    diabetes = `Diagnosed diabetes among adults`,
    chd = `Coronary heart disease among adults`,
    state_name = StateDesc
  ) %>%
  mutate(state_name = str_to_lower(state_name))

# Load state geometry
states_sf <- states(cb = TRUE) %>%
  st_as_sf() %>%
  filter(!STUSPS %in% c("HI", "AK", "PR")) %>%
  mutate(
    state_name = str_to_lower(NAME),
    STUSPS = as.character(STUSPS)
  )

# Merge geometry and summary
map_data <- left_join(states_sf, state_summary, by = "state_name") %>%
  drop_na(high_bp, diabetes, chd) %>%
  mutate(
    r = rescale(high_bp),
    g = rescale(diabetes),
    b = rescale(chd),
    rgb_fill = rgb(r, g, b)
  )

# Add label coordinates
state_labels <- map_data
centroids <- st_centroid(state_labels$geometry)
coords <- st_coordinates(centroids)
state_labels$lon <- coords[, 1]
state_labels$lat <- coords[, 2]

# High Blood Pressure Map
hbp <- ggplot(map_data) +
  geom_sf(aes(fill = high_bp), color = "white") +
  scale_fill_viridis_c(option = "C", name = "Prevalence") +
  geom_text(data = state_labels, aes(x = lon, y = lat, label = STUSPS),
            color = "black", size = 2) +
  labs(title = "High Blood Pressure Among Adults by State") +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 50)) +
  theme_minimal()

# Diagnosed Diabetes Map
dd <- ggplot(map_data) +
  geom_sf(aes(fill = diabetes), color = "white") +
  scale_fill_viridis_c(option = "C", name = "Prevalence") +
  geom_text(data = state_labels, aes(x = lon, y = lat, label = STUSPS),
            color = "black", size = 2) +
  labs(title = "Diagnosed Diabetes Among Adults by State") +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 50)) +
  theme_minimal()

# Coronary Heart Disease Map
chd <- ggplot(map_data) +
  geom_sf(aes(fill = chd), color = "white") +
  scale_fill_viridis_c(option = "C", name = "Prevalence") +
  geom_text(data = state_labels, aes(x = lon, y = lat, label = STUSPS),
            color = "black", size = 2) +
  labs(title = "Coronary Heart Disease Among Adults by State") +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 50)) +
  theme_minimal()

# Print all maps
print(hbp)
print(dd)
print(chd)

```

***Figure 1: Choropleth mapped analysis of cardiovascular disease prevalence in the U.S.***

#### **1.2 Rural vs. Urban Disparities of Cardiovascular Disease in State with the Highest Prevalence**

The data reveals that the state with the highest prevalence of cardiovascular disease in the USA is Mississippi. The boxplot analysis shows a clear disparity in the chronic disease prevalence between urban and rural populations. This analysis underscores a systematic rural disadvantage in health outcome within the state evidenced by significant difference prevalence of p\< 0.05 in all three conditions. This outcome suggests that systematized barriers such as access of preventative care and socioeconomic status are key contributors to the disease prevalence.

```{r}
library(readxl)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(ggpubr)

#Loading and cleaning rural-urban classification data
rural_urban <- read_excel("PopulationEstimates.xlsx", sheet = "Population", skip = 4) %>%
  rename(
    FIPS = FIPStxt,
    State = State,
    CountyName = Area_Name,
    RUCC_2023 = `Rural_Urban_Continuum_Code_2023`
  ) %>%
  select(FIPS, State, CountyName, RUCC_2023) %>%
  mutate(FIPS = str_pad(as.character(FIPS), 5, pad = "0")) %>%
  filter(!is.na(RUCC_2023))

#Cleaning for joins
normalize_county <- function(name) {
  name %>%
    tolower() %>%
    str_replace_all("(?<!saint)\\bst\\.?\\b", "saint") %>%
    str_remove(" county$") %>%
    str_remove(" parish$") %>%
    str_squish()
}

places_clean <- places_clean %>%
  mutate(
    StateDesc = tolower(StateDesc),
    LocationName = normalize_county(LocationName)
  )

fips_master_dedup <- fips_master_clean %>%
  mutate(
    state_name = tolower(state_name),
    county_name = normalize_county(county_name),
    FIPS = str_pad(fips, 5, pad = "0")
  ) %>%
  distinct(state_name, county_name, .keep_all = TRUE)

#Merge datasets and classifying area type
places_with_fips <- places_clean %>%
  left_join(fips_master_dedup, by = c("StateDesc" = "state_name", "LocationName" = "county_name")) %>%
  filter(!is.na(fips)) %>%
  mutate(FIPS = str_pad(as.character(fips), 5, pad = "0")) %>%
  left_join(rural_urban, by = "FIPS") %>%
  mutate(
    AreaType = case_when(
      RUCC_2023 %in% 1:3 ~ "Urban",
      RUCC_2023 %in% 4:9 ~ "Rural",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(AreaType), !is.na(Data_Value))

#Identify top state by combined prevalence of 3 conditions
combined_prevalence <- places_with_fips %>%
  filter(Measure %in% c(
    "High blood pressure among adults",
    "Diagnosed diabetes among adults",
    "Coronary heart disease among adults"
  )) %>%
  group_by(StateDesc, Measure) %>%
  summarize(mean_val = mean(Data_Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Measure, values_from = mean_val) %>%
  mutate(
    combined_score = rowMeans(across(
      c("High blood pressure among adults", 
        "Diagnosed diabetes among adults", 
        "Coronary heart disease among adults")
    ), na.rm = TRUE)
  ) %>%
  arrange(desc(combined_score))

top_state <- combined_prevalence %>% slice(1)
top_state_name <- top_state$StateDesc

#Filter for top state and simplify labels
box_data <- places_with_fips %>%
  filter(StateDesc == top_state_name) %>%
  filter(Measure %in% c(
    "High blood pressure among adults",
    "Diagnosed diabetes among adults",
    "Coronary heart disease among adults"
  )) %>%
  filter(!is.na(Data_Value), !is.na(AreaType)) %>%
  mutate(Measure = case_when(
    Measure == "High blood pressure among adults" ~ "High blood pressure",
    Measure == "Diagnosed diabetes among adults" ~ "Diagnosed diabetes",
    Measure == "Coronary heart disease among adults" ~ "Coronary heart disease",
    TRUE ~ Measure
  ))

#plotting boxplot
ggplot(box_data, aes(x = AreaType, y = Data_Value, fill = AreaType)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.9) +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 3, color = "black", fill = "white") +
  facet_wrap(~ Measure, scales = "free_y") +
  stat_compare_means(
    method = "t.test",
    label = "p.signif",     # shows stars (*, **, ***)
    label.y.npc = "top",    # position label at top of boxplot
    size = 5
  ) +
  scale_fill_manual(values = c("Urban" = "#1f77b4", "Rural" = "#ff7f0e")) +
  labs(
    title = paste("Urban vs Rural Cardiovascular Disease Prevalence in", str_to_title(top_state_name)),
    x = "Area Type",
    y = "Prevalence (%)",
    fill = "Area Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 11),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 11),
    legend.position = "none",
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5)  # center title
  )
```

***Figure 2: Urban vs Rural Prevalence of Cardiovascular disease in Mississippi***

#### **1.3 Cardiovascular Disease Vs Median Household Income (2022)**

This analysis as seen in Figure 3 reveals a coherent relationship between median household income and the prevalence of cardiovascular disease across U.S. counties. Figure 3 shows that low-income counties have higher burden which shows that heart disease prevalence is above 10% for counties with income below \$60,000. This result suggests socioeconomic factors significantly influence public health outcomes and this reinforces the need for upstream health interventions and community-level monitoring to mitigate long-term disease progression in vulnerable populations.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(scales)

#Dataset
income_data <- read_excel("Unemployment2023.xlsx", sheet = "Unemployment Med HH Income", skip = 4) %>%
  select(FIPS_Code, State, Area_Name, Median_Household_Income_2022) %>%
  filter(!is.na(FIPS_Code), !is.na(Median_Household_Income_2022)) %>%
  mutate(FIPS = str_pad(as.character(FIPS_Code), 5, pad = "0"))

normalize_county <- function(name) {
  name %>%
    tolower() %>%
    str_replace_all("(?<!saint)\\bst\\.?\\b", "saint") %>%
    str_remove(" county$") %>%
    str_remove(" parish$") %>%
    str_squish()
}

places_filtered <- places_clean %>%
  filter(Measure %in% c(
    "High blood pressure among adults",
    "Diagnosed diabetes among adults",
    "Coronary heart disease among adults"
  )) %>%
  mutate(
    StateDesc = tolower(StateDesc),
    LocationName = normalize_county(LocationName)
  )

fips_master_dedup <- fips_master_clean %>%
  mutate(
    state_name = tolower(state_name),
    county_name = normalize_county(county_name),
    FIPS = str_pad(as.character(fips), 5, pad = "0")
  ) %>%
  distinct(state_name, county_name, .keep_all = TRUE)

#merging dataset
places_with_fips <- places_filtered %>%
  left_join(fips_master_dedup, by = c("StateDesc" = "state_name", "LocationName" = "county_name"))

places_income <- places_with_fips %>%
  filter(!is.na(fips)) %>%
  mutate(FIPS = str_pad(as.character(fips), 5, pad = "0")) %>%
  inner_join(income_data, by = "FIPS") %>%
  select(StateDesc, LocationName, FIPS, Measure, Data_Value, Median_Household_Income_2022)

#data modification for plotting
places_income <- places_income %>%
  mutate(Measure = case_when(
    Measure == "High blood pressure among adults" ~ "High blood pressure",
    Measure == "Diagnosed diabetes among adults" ~ "Diagnosed diabetes",
    Measure == "Coronary heart disease among adults" ~ "Coronary heart disease",
    TRUE ~ Measure
  ))

#scatterplot at county level
ggplot(places_income, aes(x = Median_Household_Income_2022, y = Data_Value, color = Measure)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, aes(color = Measure), linetype = "dashed") +
  labs(
    title = "Economic Disparities and Chronic Disease Prevalence in U.S. Counties",
    x = "Median Household Income (USD)",
    y = "Prevalence (%)",
    color = "Condition"
  ) +
  scale_x_continuous(labels = dollar_format(scale = 1)) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 11, face = "bold", hjust = 0.5),
    legend.position = "top",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9)
  )


```

#### **1.4 Heart Disease Prevalence Vs Median Household Income in the States with the Highest and Lowest Prevalence of Chronic Disease (2022)**

The comparative dataset of state with the highest and lowest indices of chronic disease and income gap using data from the U.S Department of Agriculture’s Economic Research Service Median Household Income reveals that Counties in Mississippi consistently report high chronic disease prevalence and lower household income compared to low-burden District of Columbia. Figure 4 illustrates that counties with lower median household incomes tend to have higher prevalence of coronary heart disease. This trend highlights potential care access challenges and lifestyle risk factors tied to poverty.

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(scales)
library(ggrepel)

#Dataset and cleaning
income_data <- read_excel("Unemployment2023.xlsx", sheet = "Unemployment Med HH Income", skip = 4) %>%
  select(FIPS_Code, State, Area_Name, Median_Household_Income_2022) %>%
  filter(!is.na(FIPS_Code), !is.na(Median_Household_Income_2022)) %>%
  mutate(FIPS = str_pad(as.character(FIPS_Code), 5, pad = "0"))

normalize_county <- function(name) {
  name %>%
    tolower() %>%
    str_replace_all("(?<!saint)\\bst\\.?\\b", "saint") %>%
    str_remove(" county$| parish$") %>%
    str_squish()
}

places_filtered <- places_clean %>%
  filter(Measure %in% c("Diagnosed diabetes among adults", "Coronary heart disease among adults")) %>%
  mutate(
    StateDesc = tolower(StateDesc),
    LocationName = normalize_county(LocationName)
  )

fips_master_clean <- fips_master_clean %>%
  mutate(
    state_name = tolower(state_name),
    county_name = normalize_county(county_name),
    FIPS = str_pad(as.character(fips), 5, pad = "0")
  ) %>%
  distinct(state_name, county_name, .keep_all = TRUE)

#merging dataset
places_with_fips <- places_filtered %>%
  left_join(fips_master_clean, by = c("StateDesc" = "state_name", "LocationName" = "county_name"))

places_income <- places_with_fips %>%
  filter(!is.na(fips)) %>%
  mutate(FIPS = str_pad(as.character(fips), 5, pad = "0")) %>%
  inner_join(income_data, by = "FIPS") %>%
  select(StateDesc, LocationName, FIPS, Measure, Data_Value, Median_Household_Income_2022) %>%
  mutate(Measure = case_when(
    Measure == "Diagnosed diabetes among adults" ~ "Diagnosed diabetes",
    Measure == "Coronary heart disease among adults" ~ "Coronary heart disease"
  ))

#Identify lowest-prevalence state
lowest_prev_state <- places_income %>%
  group_by(StateDesc, Measure) %>%
  summarize(mean_prev = mean(Data_Value, na.rm = TRUE), .groups = "drop") %>%
  group_by(StateDesc) %>%
  summarize(overall_mean = mean(mean_prev), .groups = "drop") %>%
  arrange(overall_mean) %>%
  slice(1) %>%
  pull(StateDesc)

#identified as highest prevalence state so Filter to Mississippi and comparison state
comparison_data <- places_income %>%
  filter(StateDesc %in% c("mississippi", lowest_prev_state)) %>%
  mutate(StateLabel = str_to_title(StateDesc))

extreme_points <- comparison_data %>%
  group_by(Measure) %>%
  filter(Data_Value == max(Data_Value) | Median_Household_Income_2022 == min(Median_Household_Income_2022)) %>%
  ungroup()

#Plotting diagram
ggplot(comparison_data, aes(
  x = Median_Household_Income_2022,
  y = Data_Value,
  color = StateLabel,
  shape = Measure
)) +
  geom_point(size = 3, alpha = 0.75) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_text_repel(
    data = subset(comparison_data, StateLabel == "District Of Columbia"),
    aes(label = "DC"),
    size = 3.2,
    show.legend = FALSE
  ) +
  geom_text_repel(
    data = extreme_points,
    aes(label = LocationName),
    size = 3.2,
    max.overlaps = 10,
    show.legend = FALSE
  ) +
  labs(
    title = "Contrasting Diabetes and Coronary Heart Disease Burden and Income Levels",
    subtitle = paste("Mississippi vs.", str_to_title(lowest_prev_state)),
    x = "Median Household Income (USD)",
    y = "Prevalence (%)",
    color = "State",
    shape = "Condition"
  ) +
  scale_x_continuous(labels = dollar_format()) +
  guides(
    color = guide_legend(order = 1),
    shape = guide_legend(order = 2)
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12,hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom",
    legend.box = "vertical",               # Stack color (State) above shape (Condition)
    legend.box.just = "center",
    legend.title = element_text(face = "bold"),
    legend.spacing.y = unit(0.001,"cm")    # 🔍 minimal space between the two rows
  )

```

### Conclusions

This report reveals clear patterns in the geographic distribution and co-prevalence of cardiovascular risks across U.S. counties. This project discovered strong correlations between income levels and the prevalence of cardiovascular diseases underscoring clusters of elevated risk in economically disadvantaged counties, particularly in the South and Midwest. Our analysis highlights stark disparities that point to urgent needs for targeted public health interventions. These would include.

1.  Investing in primary care.

2.  Deployment of mobile health clinics.

3.  Establishing and funding local partnerships and non-profits for lifestyle coaching and health monitoring and education programs.

Our data-driven approach has uncovered actionable insights into cardiovascular health disparities, demonstrating our ability to deliver impactful, policy-relevant findings. As a committed consulting firm, we bring high-quality analytical support and impactful insights.  

### Appendices

#### Citation

1.  Centers for Disease Control and Prevention. Heart Disease Facts. U.S. Department of Health & Human Services, 15 Feb. 2024, [https://stacks.cdc.gov/view/cdc/6321. Accessed 30 May 2025](https://stacks.cdc.gov/view/cdc/6321.%20Accessed%2030%20May%202025).

2.  Centers for Disease Control and Prevention. PLACES: Local Data for Better Health, County Data 2024 Release. U.S. Department of Health & Human Services, 30 Apr. 2024, <https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-County-Data-20/swc5-untb/about_data>.  Accessed  23 April 2025.

3.  Healy, Kieran. county_fips_master.csv. GitHub, <https://github.com/kjhealy/fips-codes/blob/master/county_fips_master.csv>.  Accessed 27 April 2025.

4.  U.S. Department of Agriculture, Economic Research Service. County-Level Data Sets: Download Data. 29 Feb. 2024, <https://www.ers.usda.gov/data-products/county-level-data-sets/county-level-data-sets-download-data>.  Accessed 23 April 2025.
